---
title: "02 Exploratory analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(tidymodels)
library(modeltime)
library(lubridate)
library(timetk)

data_url <- "https://raw.githubusercontent.com/DiabbZegpi/AI-eval2/main/datos_precios_train.csv"
submission_url <- "https://raw.githubusercontent.com/DiabbZegpi/AI-eval2/main/datos_precio_prediction.csv"

data <- read_csv(data_url) |> 
  mutate(Fecha = dmy(Fecha)) |> 
  arrange(Fecha)

submission_sample <- read_csv(submission_url) |> 
  mutate(Fecha = dmy(Fecha)) |> 
  arrange(Fecha)
```

## *Plot configuration*

```{r}
knitr::opts_chunk$set(fig.width = 11)
theme_set(theme_minimal())
theme_update(
  text = element_text(size = 14),
  plot.title = element_text(hjust = 0.5, face = "bold"),
  plot.subtitle = element_text(hjust = 0.5),
  panel.grid.minor = element_blank()
)
```

## *Data partition*

```{r}
splits <- initial_time_split(data, prop = 3/4)
training_data <- training(splits)
testing_data <- testing(splits)
folds <- time_series_cv(
  training_data,
  date_var = Fecha,
  initial = 60, 
  skip = 5, 
  assess = 20
)
```

```{r}
plot_ts <- function(data, ribbon = TRUE) {
 date_var <- map_lgl(data, is.Date) |> which() |> names()
  data_points <- nrow(data)
  range_dates <- data |> pull({{ date_var }}) |> range()
  ttl_lab <- sprintf("Training data: %i observations", data_points)
  subttl_lab <- sprintf("From %s to %s", range_dates[1], range_dates[2])
  
  p <- ggplot(data, aes_string(x = date_var)) +
    geom_line(aes(y = Ultimo)) +
    labs(title = ttl_lab, subtitle = subttl_lab,
         x = "Date", y = "Stock price")
  
  if (ribbon) p + geom_ribbon(aes(ymax = Maximo, ymin = Minimo), fill = "cyan4", alpha = 0.5) else p
}

plot_ts(training_data)
```

The time series has a frequency of 5 days, with no evident trend nor seasonality. It is clear that the series aren't stationary.

```{r}
plot_ts(training_data |> filter(Fecha < "2016-01-01")) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")
```

Which order of differentiation would stationarize the series characteristics? 

```{r}
library(patchwork)
p1 <- plot_ts(training_data |> mutate(Ultimo = Ultimo - lag(Ultimo)), ribbon = FALSE)
p2 <- plot_ts(training_data |> mutate(Ultimo = Ultimo - lag(Ultimo, 5)), ribbon = FALSE)
p3 <- plot_ts(training_data |> mutate(Ultimo = Ultimo - lag(Ultimo) - lag(Ultimo, 5)), ribbon = FALSE)
p4 <- plot_ts(training_data |> mutate(Ultimo = Ultimo - lag(Ultimo) - lag(Ultimo)), ribbon = FALSE)

p1 + p2 + p3 + p4 + plot_layout(ncol = 2) + plot_annotation(tag_levels = "I")
```

Only 1st order non-seasonal differentiation is necessary to make the series staationary.

## Visualizing cross-validation plan

```{r}
folds |> 
  tk_time_series_cv_plan() |> 
  filter(parse_number(.id) <= 20) |> 
  plot_ts(ribbon = FALSE) +
  aes(color = .key) +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap(~.id) +
  guides(color = "none")
```

Apparently, for the first year of data, stock prices go up for the second half of the year, while oscillating at the beginning of the period.

## *Fit GLUON deep learning models*

```{r}
library(modeltime.gluonts)

gluon_spec <- deep_ar(
  id = "id",
  freq = "B",
  prediction_length = 264, 
  lookback_length = NULL, 
  cell_type = "lstm",
  num_layers = 3, 
  num_cells = 40,
  dropout = 0.4,
  epochs = 20,
  batch_size = 16,
  learn_rate = 0.001
) |> 
  set_engine("gluonts_deepar")

gluon_fit_ultimo <- gluon_spec |> 
  fit(Ultimo ~ Fecha + id, data = training_data |> mutate(id = "id"))

gluon_fit_maximo <- gluon_spec |> 
  fit(Maximo ~ Fecha + id, data = training_data |> mutate(id = "id"))
```

```{r}
submission_sample |> pivot_longer(-Fecha, names_to = "id", values_to = "stock")

gluon_preds <- gluon_fit |> 
  modeltime_table() |> 
  modeltime_calibrate(gluon_test) |> 
  modeltime_forecast(
    new_data = submission_sample |> 
      pivot_longer(-Fecha, names_to = "id", values_to = "stock"),
    actual_data = gluon_data,
    keep_data = TRUE
  ) 

preds_to_submit <- gluon_preds |> 
  filter(.key == "prediction") |> 
  select(Fecha, .value, id) |> 
  pivot_wider(names_from = id, values_from = .value) 

preds_to_submit$Fecha <- format(preds_to_submit$Fecha, "%d-%m-%Y")
filename <- paste0("../Submissions/", Sys.Date(), "_GLUON_AR.csv")
write_csv(preds_to_submit, file = filename)
```


